\renewcommand{\appendixname}{\textcolor{white}{.}}
\renewcommand{\thefigure}{S\arabic{figure}} \setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}} \setcounter{table}{0}
\setcounter{page}{1} <!-- to reset page numbering -->

# Supplemental Material

\noindent
**Supplemental Tables**  

```{r icc-table, results="asis"}
# ICCs = proportion of the total variation explained by the respective blocking factor
# https://stats.stackexchange.com/questions/18088/intraclass-correlation-icc-for-an-interaction
outcomes <- c("agree", "con", "extra", "neur", "open", "swls")
datasets <- c("lissanalysis_parents", "lissanalysis_nonparents", 
              "hrsanalysis_parents", "hrsanalysis_nonparents")
icc_list <- as.data.frame(rbind(icc_liss_parents_pid = rep(NA, 6), icc_liss_parents_hid = rep(NA, 6), 
                                icc_liss_parents_pidhid = rep(NA, 6), icc_liss_nonparents_pid = rep(NA, 6),
                                icc_liss_nonparents_hid = rep(NA, 6), 
                                icc_liss_nonparents_pidhid = rep(NA, 6), icc_hrs_parents_pid = rep(NA, 6), 
                                icc_hrs_parents_hid = rep(NA, 6), icc_hrs_parents_pidhid = rep(NA, 6), 
                                icc_hrs_nonparents_pid = rep(NA, 6), icc_hrs_nonparents_hid = rep(NA, 6),
                                icc_hrs_nonparents_pidhid = rep(NA, 6)))
colnames(icc_list) <- outcomes

for (i in 1:length(outcomes)){
  outcome = outcomes[i]
  for (j in 1:length(datasets)){
    dataset = datasets[j]
    pos = seq(from = 1, to = 10, by = 3)[j]
    model <- lme4::lmer(get(outcome) ~ 1 + (1 | pid) + (1 | hid), REML = FALSE, data = get(dataset))
    var <- as.data.frame(summary(model)$varcor)
    icc_pid <- var[1,4] / (var[1,4] + var[2,4] + var[3,4])
    icc_hid <- var[2,4] / (var[1,4] + var[2,4] + var[3,4])
    icc_pidhid <- (var[1,4] + var[2,4]) / (var[1,4] + var[2,4] + var[3,4])
    icc_list[pos, i] <- icc_pid # rows = datasets, columns = outcomes
    icc_list[pos + 1, i] <- icc_hid
    icc_list[pos + 2, i] <- icc_pidhid
  }
}
# singular fir for two models where the icc_hid is estimated too close to zero
# For these models, the icc_pid is the same compared to the model with only 
# the (1 | pid) random effect (which fits without trouble)

rownames(icc_list) <- c("$ICC_{pid}$ \\textcolor{white}{LP}", "$ICC_{hid}$ \\textcolor{white}{LP}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{LP}", 
                        "$ICC_{pid}$ \\textcolor{white}{LN}", "$ICC_{hid}$ \\textcolor{white}{LN}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{LN}", 
                        "$ICC_{pid}$ \\textcolor{white}{HP}", "$ICC_{hid}$ \\textcolor{white}{HP}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{HP}", 
                        "$ICC_{pid}$ \\textcolor{white}{HN}", "$ICC_{hid}$ \\textcolor{white}{HN}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{HN}")
colnames(icc_list) <- c("\\textcolor{white}{xxx}A\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}C\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}E\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}N\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}O\\textcolor{white}{xxx}", 
                        "\\textcolor{white}{xxx}LS\\textcolor{white}{xxx}") # added some whitespace to adjust table width

apa_table(
  icc_list
  , caption = "Intra-Class Correlations"
  , note = "A = agreeableness, C = conscientiousness, E = extraversion, N = neuroticism, O = openness, LS = life satisfaction. Intra-class correlations are the proportion of total variation that is explained by the respective blocking factor. $ICC_{pid}$ is the proportion of total variance explained by nesting in respondents which corresponds to the correlation between two randomly selected observations from the same respondent. $ICC_{hid}$ is the proportion of total variance explained by nesting in households which corresponds to the correlation between two randomly selected observations from the same household. $ICC_{pid/hid}$ is the proportion of total variance explained by nesting in respondents and in households which corresponds to the correlation between two randomly selected observations from the same respondent and the same household." 
  , align = c("l", "c", "c", "c", "c", "c", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  , stub_indents = list("LISS: Parent controls" = 1:3, "LISS: Nonparent controls" = 4:6,
                        "HRS: Parent controls" = 7:9, "HRS: Nonparent controls" = 10:12)
  , escape = F
  , landscape = F
  , placement = "h"
)
```

```{r draw-longitudinal-obs-info, include=FALSE}
# show overview of sample size over time in the final analysis samples
hrsanalysis_parents <- hrsanalysis_parents %>% mutate(female = gender - 1)
hrsanalysis_nonparents <- hrsanalysis_nonparents %>% mutate(female = gender - 1)

sample_time_draw <- function(x) { 
  x %>% 
    group_by(grandparent, time) %>% summarise(n=n(), f = mean(female)*100) %>%
    mutate(f = format(round(f, digits=2), nsmall = 2)) %>% # fix at 2 decimal places
    pivot_wider(names_from = time, values_from = -c(time, grandparent)) %>% 
    mutate_if(is.numeric, as.character) %>% # transform num to chr
    pivot_longer(cols = -grandparent, names_to = c("var", ".value"), names_pattern = "(.)_(.*)") }
  
list_sample_time <- list(lissanalysis_parents, lissanalysis_nonparents, 
                         hrsanalysis_parents, hrsanalysis_nonparents) %>%
  lapply(sample_time_draw)
names(list_sample_time) <- c("sample_time_liss_p", "sample_time_liss_np", 
                             "sample_time_hrs_p", "sample_time_hrs_np")
list2env(list_sample_time, .GlobalEnv)
rm(list_sample_time)

# we need grandparent obs only once per sample
sample_time_liss_p <- sample_time_liss_p %>% arrange(desc(grandparent))
sample_time_liss_np <- sample_time_liss_np %>% filter(grandparent==0)
sample_time_hrs_p <- sample_time_hrs_p %>% arrange(desc(grandparent))
sample_time_hrs_np <- sample_time_hrs_np %>% filter(grandparent==0)
```

(ref:piecewise-coding-scheme-cap) Longitudinal sample size in the analysis samples and coding scheme for the piecewise regression coefficients

(ref:piecewise-coding-scheme-note) obs. = observations. $time=0$ marks the first year where the transition to grandparenthood has been reported. The number of grandparent participants is $N_{LISS}=$ 250 and $N_{HRS}=$ 846. 

```{r piecewise-coding-scheme, results="asis", warning=FALSE, message=FALSE}
sample_time_liss <- as.data.frame(rbind(sample_time_liss_p, sample_time_liss_np))
sample_time_liss <- sample_time_liss[,-c(1,2)]
sample_time_hrs <- as.data.frame(rbind(sample_time_hrs_p, sample_time_hrs_np))
sample_time_hrs <- sample_time_hrs[,-c(1,2)]
sample_time_hrs <- sample_time_hrs %>% add_column(`-5` = "", `-3` = "", `-1` = "", 
                                                  `1` = "", `3` = "", `5` = "")

# Visualize the coding scheme for the piecewise regression coefficients
coding_scheme_liss <- rbind(c("0", "1", "2", "3", "4", "5", "5", "5", "5", "5", "5", "5", "5"),
                       c("0", "0", "0", "0", "0", "0", "1", "2", "3", "4", "5", "6", "7"),
                       c("0", "0", "0", "0", "0", "0", "1", "1", "1", "1", "1", "1", "1"))
colnames(coding_scheme_liss) <- c("-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")

coding_scheme_hrs <- rbind(c("0", "", "1", "", "2", "", "2", "", "2", "", "2", "", "2"),
                           c("0", "", "0", "", "0", "", "1", "", "2", "", "3", "", "4"),
                           c("0", "", "0", "", "0", "", "1", "", "1", "", "1", "", "1"))
colnames(coding_scheme_hrs) <- c("-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")

# combine both
sample_time_and_coding <- rbind(sample_time_liss, coding_scheme_liss,
                                sample_time_hrs, coding_scheme_hrs)

# no duplicate rownames allowed :(
rownames(sample_time_and_coding) <- c("Grandparents: obs. \\textcolor{white}{L}",
                          "Grandparents: \\% women \\textcolor{white}{L}", 
                          "Parent controls: obs. \\textcolor{white}{L}", 
                          "Parent controls: \\% women \\textcolor{white}{L}", 
                          "Nonparent controls: obs. \\textcolor{white}{L}", 
                          "Nonparent controls: \\% women \\textcolor{white}{L}", 
                          "Before-slope \\textcolor{white}{L}", "After-slope \\textcolor{white}{L}",
                          "Jump \\textcolor{white}{L}", 
                          "Grandparents: obs. \\textcolor{white}{H}",
                          "Grandparents: \\% women \\textcolor{white}{H}", 
                          "Parent controls: obs. \\textcolor{white}{H}", 
                          "Parent controls: \\% women \\textcolor{white}{H}", 
                          "Nonparent controls: obs. \\textcolor{white}{H}", 
                          "Nonparent controls: \\% women \\textcolor{white}{H}",
                          "Before-slope \\textcolor{white}{H}", "After-slope \\textcolor{white}{H}",
                          "Jump \\textcolor{white}{H}")

apa_table(
  sample_time_and_coding
  , caption = "(ref:piecewise-coding-scheme-cap)"
  , note = "(ref:piecewise-coding-scheme-note)"
  , align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
  , col.names = c("", "-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")
  , col_spanners = list("Pre-transition years" = c(2,7), "Post-transition years" = c(8,14))
  , stub_indents = list("LISS: Analysis samples" = 1:6, "LISS: Coding scheme" = 7:9, 
                        "HRS: Analysis samples" = 10:15, "HRS: Coding scheme" = 16:18)
  , format.args = list(gt1 = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T),
                       digits = c(0,0,0,0,0,0,0,0,0,0,0,0,0),
                       zero = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T))
  , escape = F
  , landscape = T
  , placement = "ht"
  , font_size = "small"
)
```

```{r draw-descriptives, include=FALSE}
descriptives_draw <- function(x) { 
  x %>% 
  group_by(grandparent, time) %>% summarize(a.M = mean(agree, na.rm = T), 
                                            a.SD = sd(agree, na.rm = T),
                                            c.M = mean(con, na.rm = T), 
                                            c.SD = sd(con, na.rm = T),
                                            e.M = mean(extra, na.rm = T), 
                                            e.SD = sd(extra, na.rm = T),
                                            n.M = mean(neur, na.rm = T), 
                                            n.SD = sd(neur, na.rm = T),
                                            o.M = mean(open, na.rm = T), 
                                            o.SD = sd(open, na.rm = T),
                                            l.M = mean(swls, na.rm = T), 
                                            l.SD = sd(swls, na.rm = T)) %>%
  ungroup() %>% 
  mutate_at(.vars = vars(-c(grandparent, time)), 
            ~format(round(., digits=2), nsmall = 2)) %>% # fix at 2 decimal places
  pivot_wider(names_from = time, values_from = -c(time, grandparent)) %>% 
  mutate_if(is.numeric, as.character) %>% # transform num to chr
  mutate_at(.vars = vars(contains("SD")), 
            ~paste0("(", ., ")")) %>% # add parantheses for SD
  pivot_longer(cols = -grandparent, names_to = c("var", ".value"), names_pattern = "(.*)_(.*)") }

list_descriptives <- list(lissanalysis_parents, lissanalysis_nonparents, 
                         hrsanalysis_parents, hrsanalysis_nonparents) %>%
  lapply(descriptives_draw)
names(list_descriptives) <- c("descriptives_liss_p", "descriptives_liss_np", 
                             "descriptives_hrs_p", "descriptives_hrs_np")
list2env(list_descriptives, .GlobalEnv)
rm(list_descriptives)

# we need grandparent obs only once per sample
descriptives_liss_p <- descriptives_liss_p %>% arrange(desc(grandparent))
descriptives_liss_np <- descriptives_liss_np %>% filter(grandparent==0)
descriptives_hrs_p <- descriptives_hrs_p %>% arrange(desc(grandparent))
descriptives_hrs_np <- descriptives_hrs_np %>% filter(grandparent==0)

descriptives_liss <- as.data.frame(rbind(descriptives_liss_p, descriptives_liss_np))
descriptives_liss <- descriptives_liss %>% 
  slice(1:2, 13:14, 25:26, 3:4, 15:16, 27:28, 5:6, 17:18, 29:30, 7:8, 19:20, 31:32, 
        9:10, 21:22, 33:34, 11:12, 23:24, 35:36)
descriptives_liss <- descriptives_liss[,-c(1,2)]
descriptives_hrs <- as.data.frame(rbind(descriptives_hrs_p, descriptives_hrs_np))
descriptives_hrs <- descriptives_hrs %>% 
  slice(1:2, 13:14, 25:26, 3:4, 15:16, 27:28, 5:6, 17:18, 29:30, 7:8, 19:20, 31:32, 
        9:10, 21:22, 33:34, 11:12, 23:24, 35:36)
descriptives_hrs <- descriptives_hrs[,-c(1,2)]
descriptives_hrs <- descriptives_hrs %>% add_column(`-5` = "", `-3` = "", `-1` = "", 
                                                  `1` = "", `3` = "", `5` = "") %>% 
  select(`-6`, `-5`, `-4`, `-3`, `-2`, `-1`, `0`, `1`, `2`, `3`, `4`, `5`, `6`)
```

(ref:descriptives-liss-cap) Mean and Standard Deviation of the Big Five and Life Satisfaction over Time in the LISS Panel

(ref:descriptives-note) Standard deviation shown in brackets; $time=0$ marks the first year where the transition to grandparenthood has been reported.

```{r descriptives-liss, results="asis", warning=FALSE, message=FALSE}
# no duplicate rownames allowed :(
rownames(descriptives_liss) <- c("Grandparents \\textcolor{white}{A}",
                          "\\textcolor{white}{Ag}", 
                          "Parent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{Ap}", 
                          "Nonparent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{An}", 
                          "Grandparents \\textcolor{white}{C}",
                          "\\textcolor{white}{Cg}", 
                          "Parent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cp}", 
                          "Nonparent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cn}", 
                          "Grandparents \\textcolor{white}{E}",
                          "\\textcolor{white}{Eg}", 
                          "Parent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{Ep}", 
                          "Nonparent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{En}", 
                          "Grandparents \\textcolor{white}{N}",
                          "\\textcolor{white}{Ng}", 
                          "Parent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Np}", 
                          "Nonparent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Nn}", 
                          "Grandparents \\textcolor{white}{O}",
                          "\\textcolor{white}{Og}", 
                          "Parent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{Op}", 
                          "Nonparent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{On}", 
                          "Grandparents \\textcolor{white}{L}",
                          "\\textcolor{white}{Lg}", 
                          "Parent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Lp}", 
                          "Nonparent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Ln}")

apa_table(
  descriptives_liss
  , caption = "(ref:descriptives-liss-cap)"
  , note = "(ref:descriptives-note)"
  , align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
  , col.names = c("", "-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")
  , col_spanners = list("Pre-transition years" = c(2,7), "Post-transition years" = c(8,14))
  , stub_indents = list("Agreeableness" = 1:6, "Conscientiousness" = 7:12, 
                        "Extraversion" = 13:18, "Neuroticism" = 19:24,
                        "Openness" = 25:30, "Life satisfaction" = 31:36)
  , format.args = list(gt1 = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T),
                       digits = c(0,0,0,0,0,0,0,0,0,0,0,0,0),
                       zero = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T))
  , escape = F
  , landscape = T
  , placement = "ht"
  , font_size = "small"
)
```

(ref:descriptives-hrs-cap) Mean and Standard Deviation of the Big Five and Life Satisfaction over Time in the HRS

```{r descriptives-hrs, results="asis", warning=FALSE, message=FALSE}
# no duplicate rownames allowed :(
rownames(descriptives_hrs) <- c("Grandparents \\textcolor{white}{A}",
                          "\\textcolor{white}{Ag}", 
                          "Parent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{Ap}", 
                          "Nonparent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{An}", 
                          "Grandparents \\textcolor{white}{C}",
                          "\\textcolor{white}{Cg}", 
                          "Parent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cp}", 
                          "Nonparent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cn}", 
                          "Grandparents \\textcolor{white}{E}",
                          "\\textcolor{white}{Eg}", 
                          "Parent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{Ep}", 
                          "Nonparent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{En}", 
                          "Grandparents \\textcolor{white}{N}",
                          "\\textcolor{white}{Ng}", 
                          "Parent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Np}", 
                          "Nonparent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Nn}", 
                          "Grandparents \\textcolor{white}{O}",
                          "\\textcolor{white}{Og}", 
                          "Parent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{Op}", 
                          "Nonparent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{On}", 
                          "Grandparents \\textcolor{white}{L}",
                          "\\textcolor{white}{Lg}", 
                          "Parent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Lp}", 
                          "Nonparent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Ln}")

apa_table(
  descriptives_hrs
  , caption = "(ref:descriptives-hrs-cap)"
  , note = "(ref:descriptives-note)"
  , align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
  , col.names = c("", "-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")
  , col_spanners = list("Pre-transition years" = c(2,7), "Post-transition years" = c(8,14))
  , stub_indents = list("Agreeableness" = 1:6, "Conscientiousness" = 7:12, 
                        "Extraversion" = 13:18, "Neuroticism" = 19:24,
                        "Openness" = 25:30, "Life satisfaction" = 31:36)
  , format.args = list(gt1 = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T),
                       digits = c(0,0,0,0,0,0,0,0,0,0,0,0,0),
                       zero = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T))
  , escape = F
  , landscape = T
  , placement = "ht"
  , font_size = "small"
)
```

(ref:stddiffmeans-balance-liss-cap) Standardized Difference in Means for Covariates Used in Propensity Score Matching and the Propensity Score in the LISS panel 

(ref:stddiffmeans-balance-note) PSM = propensity score matching, ref. = reference category, f. = female, m. = male, NA = covariate not used in this sample. The standardized difference in means between the grandparent and the two control groups (parent and nonparent) was computed by $(\bar{x}_{gp}-\bar{x}_{c})/ (\hat\sigma_{gp})$. Rules of thumb say that this measure should ideally be below $.25$ [@stuartMatchingMethodsCausal2010] or below $.10$ [@austinIntroductionPropensityScore2011].  

```{r stddiffmeans-balance-liss, results="asis"}
# Standardized differences between means
liss_balance_df_rownames <- liss_balance_df$Covariate
liss_balance_df <- liss_balance_df[-c(1)]
rownames(liss_balance_df) <- liss_balance_df_rownames

apa_table(
  liss_balance_df
  , caption = "(ref:stddiffmeans-balance-liss-cap)"
  , note = "(ref:stddiffmeans-balance-note)"
  , align = c("l", "l", "l", "r", "r", "r", "r")
  , col.names = c("Covariate", "Description", "Raw variable", 
                  "Before PSM", "After PSM", "Before PSM", "After PSM")
  , col_spanners = list("Parent control group" = c(4,5), "Nonparent control group" = c(6,7))
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  , escape = T
  , landscape = T
  , longtable = T
  , placement = "t"
  , font_size = "footnotesize"
)
```

(ref:stddiffmeans-balance-hrs-cap) Standardized Difference in Means for Covariates Used in Propensity Score Matching and the Propensity Score in the HRS 

```{r stddiffmeans-balance-hrs, results="asis"}
# Standardized differences between means
hrs_balance_df_rownames <- hrs_balance_df$Covariate
hrs_balance_df <- hrs_balance_df[-c(1)]
rownames(hrs_balance_df) <- hrs_balance_df_rownames

apa_table(
  hrs_balance_df
  , caption = "(ref:stddiffmeans-balance-hrs-cap)"
  , note = "(ref:stddiffmeans-balance-note)"
  , align = c("l", "l", "l", "r", "r", "r", "r")
  , col.names = c("Covariate", "Description", "Raw variable", 
                  "Before PSM", "After PSM", "Before PSM", "After PSM")
  , col_spanners = list("Parent control group" = c(4,5), "Nonparent control group" = c(6,7))
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  #, stub_indents = list("Basic demographics)" = 2:14, "Economic situation" = 15:22, "Health" = 23:30,
  #                      "Family and fertility" = 31:36, "Outcomes at the time of matching" = 37:42,
  #                      "Panel status" = 43:44)
  , escape = T
  , landscape = T
  , longtable = T
  , placement = "t"
  , font_size = "footnotesize"
)
```

\newpage

\noindent
**Supplemental Figures**  

\newpage

\noindent 
**Complete Software and Session Information**  

We used `r cite_r("references-r.bib")` for data wrangling, analyses, and plots.  
The following is the output of R's *sessionInfo()* command, which shows information to aid analytic reproducibility of the analyses.

```{r session, results='asis'}

sessionInfo()

```

\newpage

\noindent 
**References**

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup
