\renewcommand{\appendixname}{\textcolor{white}{.}}

# Supplemental Material

\renewcommand{\thefigure}{S\arabic{figure}} \setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}} \setcounter{table}{0}
\setcounter{page}{1} <!-- to reset page numbering -->

\noindent
**Supplemental Tables**  

```{r icc-table, results="asis"}
# object 'icc_list' is now generated in the main .Rmd together with the other models

rownames(icc_list) <- c("$ICC_{pid}$ \\textcolor{white}{LP}", "$ICC_{hid}$ \\textcolor{white}{LP}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{LP}", 
                        "$ICC_{pid}$ \\textcolor{white}{LN}", "$ICC_{hid}$ \\textcolor{white}{LN}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{LN}", 
                        "$ICC_{pid}$ \\textcolor{white}{HP}", "$ICC_{hid}$ \\textcolor{white}{HP}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{HP}", 
                        "$ICC_{pid}$ \\textcolor{white}{HN}", "$ICC_{hid}$ \\textcolor{white}{HN}", 
                        "$ICC_{pid/hid}$ \\textcolor{white}{HN}")
colnames(icc_list) <- c("\\textcolor{white}{xxx}A\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}C\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}E\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}N\\textcolor{white}{xxx}",
                        "\\textcolor{white}{xxx}O\\textcolor{white}{xxx}", 
                        "\\textcolor{white}{xxx}LS\\textcolor{white}{xxx}") # added some whitespace to adjust table width

apa_table(
  icc_list
  , caption = "Intra-Class Correlations"
  , note = "A = agreeableness, C = conscientiousness, E = extraversion, N = neuroticism, O = openness, LS = life satisfaction. Intra-class correlations are the proportion of total variation that is explained by the respective blocking factor. $ICC_{pid}$ is the proportion of total variance explained by nesting in respondents which corresponds to the correlation between two randomly selected observations from the same respondent. $ICC_{hid}$ is the proportion of total variance explained by nesting in households which corresponds to the correlation between two randomly selected observations from the same household. $ICC_{pid/hid}$ is the proportion of total variance explained by nesting in respondents and in households which corresponds to the correlation between two randomly selected observations from the same respondent and the same household." 
  , align = c("l", "c", "c", "c", "c", "c", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  , stub_indents = list("LISS: Parent controls" = 1:3, "LISS: Nonparent controls" = 4:6,
                        "HRS: Parent controls" = 7:9, "HRS: Nonparent controls" = 10:12)
  , escape = F
  , landscape = F
  , placement = "h"
)
```

```{r draw-longitudinal-obs-info, include=FALSE}
# show overview of sample size over time in the final analysis samples
hrsanalysis_parents <- hrsanalysis_parents %>% mutate(female = gender - 1)
hrsanalysis_nonparents <- hrsanalysis_nonparents %>% mutate(female = gender - 1)

sample_time_draw <- function(x) { 
  x %>% 
    group_by(grandparent, time) %>% summarise(n=n(), f = mean(female)*100) %>%
    mutate(f = format(round(f, digits=2), nsmall = 2)) %>% # fix at 2 decimal places
    pivot_wider(names_from = time, values_from = -c(time, grandparent)) %>% 
    mutate_if(is.numeric, as.character) %>% # transform num to chr
    pivot_longer(cols = -grandparent, names_to = c("var", ".value"), names_pattern = "(.)_(.*)") }
  
list_sample_time <- list(lissanalysis_parents, lissanalysis_nonparents, 
                         hrsanalysis_parents, hrsanalysis_nonparents) %>%
  lapply(sample_time_draw)
names(list_sample_time) <- c("sample_time_liss_p", "sample_time_liss_np", 
                             "sample_time_hrs_p", "sample_time_hrs_np")
list2env(list_sample_time, .GlobalEnv)
rm(list_sample_time)

# we need grandparent obs only once per sample
sample_time_liss_p <- sample_time_liss_p %>% arrange(desc(grandparent))
sample_time_liss_np <- sample_time_liss_np %>% filter(grandparent==0)
sample_time_hrs_p <- sample_time_hrs_p %>% arrange(desc(grandparent))
sample_time_hrs_np <- sample_time_hrs_np %>% filter(grandparent==0)
```

(ref:piecewise-coding-scheme-cap) Longitudinal sample size in the analysis samples and coding scheme for the piecewise regression coefficients.

(ref:piecewise-coding-scheme-note) obs. = observations. $time=0$ marks the first year where the transition to grandparenthood has been reported. The number of grandparent participants is $N_{LISS}=$ 250 and $N_{HRS}=$ 846. 

```{r piecewise-coding-scheme, results="asis", warning=FALSE, message=FALSE}
sample_time_liss <- as.data.frame(rbind(sample_time_liss_p, sample_time_liss_np))
sample_time_liss <- sample_time_liss[,-c(1,2)]
sample_time_hrs <- as.data.frame(rbind(sample_time_hrs_p, sample_time_hrs_np))
sample_time_hrs <- sample_time_hrs[,-c(1,2)]
sample_time_hrs <- sample_time_hrs %>% add_column(`-5` = "", `-3` = "", `-1` = "", 
                                                  `1` = "", `3` = "", `5` = "")

# Visualize the coding scheme for the piecewise regression coefficients
coding_scheme_liss <- rbind(c("0", "1", "2", "3", "4", "5", "5", "5", "5", "5", "5", "5", "5"),
                       c("0", "0", "0", "0", "0", "0", "1", "2", "3", "4", "5", "6", "7"),
                       c("0", "0", "0", "0", "0", "0", "1", "1", "1", "1", "1", "1", "1"))
colnames(coding_scheme_liss) <- c("-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")

coding_scheme_hrs <- rbind(c("0", "", "1", "", "2", "", "2", "", "2", "", "2", "", "2"),
                           c("0", "", "0", "", "0", "", "1", "", "2", "", "3", "", "4"),
                           c("0", "", "0", "", "0", "", "1", "", "1", "", "1", "", "1"))
colnames(coding_scheme_hrs) <- c("-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")

# combine both
sample_time_and_coding <- rbind(sample_time_liss, coding_scheme_liss,
                                sample_time_hrs, coding_scheme_hrs)

# no duplicate rownames allowed :(
rownames(sample_time_and_coding) <- c("Grandparents: obs. \\textcolor{white}{L}",
                          "Grandparents: \\% women \\textcolor{white}{L}", 
                          "Parent controls: obs. \\textcolor{white}{L}", 
                          "Parent controls: \\% women \\textcolor{white}{L}", 
                          "Nonparent controls: obs. \\textcolor{white}{L}", 
                          "Nonparent controls: \\% women \\textcolor{white}{L}", 
                          "Before-slope \\textcolor{white}{L}", "After-slope \\textcolor{white}{L}",
                          "Jump \\textcolor{white}{L}", 
                          "Grandparents: obs. \\textcolor{white}{H}",
                          "Grandparents: \\% women \\textcolor{white}{H}", 
                          "Parent controls: obs. \\textcolor{white}{H}", 
                          "Parent controls: \\% women \\textcolor{white}{H}", 
                          "Nonparent controls: obs. \\textcolor{white}{H}", 
                          "Nonparent controls: \\% women \\textcolor{white}{H}",
                          "Before-slope \\textcolor{white}{H}", "After-slope \\textcolor{white}{H}",
                          "Jump \\textcolor{white}{H}")

apa_table(
  sample_time_and_coding
  , caption = "(ref:piecewise-coding-scheme-cap)"
  , note = "(ref:piecewise-coding-scheme-note)"
  , align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
  , col.names = c("", "-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")
  , col_spanners = list("Pre-transition years" = c(2,7), "Post-transition years" = c(8,14))
  , stub_indents = list("LISS: Analysis samples" = 1:6, "LISS: Coding scheme" = 7:9, 
                        "HRS: Analysis samples" = 10:15, "HRS: Coding scheme" = 16:18)
  , format.args = list(gt1 = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T),
                       digits = c(0,0,0,0,0,0,0,0,0,0,0,0,0),
                       zero = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T))
  , escape = F
  , landscape = T
  , placement = "ht"
  , font_size = "small"
)
```

```{r draw-descriptives, include=FALSE}
descriptives_draw <- function(x) { 
  x %>% 
  group_by(grandparent, time) %>% summarize(a.M = mean(agree, na.rm = T), 
                                            a.SD = sd(agree, na.rm = T),
                                            c.M = mean(con, na.rm = T), 
                                            c.SD = sd(con, na.rm = T),
                                            e.M = mean(extra, na.rm = T), 
                                            e.SD = sd(extra, na.rm = T),
                                            n.M = mean(neur, na.rm = T), 
                                            n.SD = sd(neur, na.rm = T),
                                            o.M = mean(open, na.rm = T), 
                                            o.SD = sd(open, na.rm = T),
                                            l.M = mean(swls, na.rm = T), 
                                            l.SD = sd(swls, na.rm = T)) %>%
  ungroup() %>% 
  mutate_at(.vars = vars(-c(grandparent, time)), 
            ~format(round(., digits=2), nsmall = 2)) %>% # fix at 2 decimal places
  pivot_wider(names_from = time, values_from = -c(time, grandparent)) %>% 
  mutate_if(is.numeric, as.character) %>% # transform num to chr
  mutate_at(.vars = vars(contains("SD")), 
            ~paste0("(", ., ")")) %>% # add parantheses for SD
  pivot_longer(cols = -grandparent, names_to = c("var", ".value"), names_pattern = "(.*)_(.*)") }

list_descriptives <- list(lissanalysis_parents, lissanalysis_nonparents, 
                         hrsanalysis_parents, hrsanalysis_nonparents) %>%
  lapply(descriptives_draw)
names(list_descriptives) <- c("descriptives_liss_p", "descriptives_liss_np", 
                             "descriptives_hrs_p", "descriptives_hrs_np")
list2env(list_descriptives, .GlobalEnv)
rm(list_descriptives)

# we need grandparent obs only once per sample
descriptives_liss_p <- descriptives_liss_p %>% arrange(desc(grandparent))
descriptives_liss_np <- descriptives_liss_np %>% filter(grandparent==0)
descriptives_hrs_p <- descriptives_hrs_p %>% arrange(desc(grandparent))
descriptives_hrs_np <- descriptives_hrs_np %>% filter(grandparent==0)

descriptives_liss <- as.data.frame(rbind(descriptives_liss_p, descriptives_liss_np))
descriptives_liss <- descriptives_liss %>% 
  slice(1:2, 13:14, 25:26, 3:4, 15:16, 27:28, 5:6, 17:18, 29:30, 7:8, 19:20, 31:32, 
        9:10, 21:22, 33:34, 11:12, 23:24, 35:36)
descriptives_liss <- descriptives_liss[,-c(1,2)]
descriptives_hrs <- as.data.frame(rbind(descriptives_hrs_p, descriptives_hrs_np))
descriptives_hrs <- descriptives_hrs %>% 
  slice(1:2, 13:14, 25:26, 3:4, 15:16, 27:28, 5:6, 17:18, 29:30, 7:8, 19:20, 31:32, 
        9:10, 21:22, 33:34, 11:12, 23:24, 35:36)
descriptives_hrs <- descriptives_hrs[,-c(1,2)]
descriptives_hrs <- descriptives_hrs %>% add_column(`-5` = "", `-3` = "", `-1` = "", 
                                                  `1` = "", `3` = "", `5` = "") %>% 
  select(`-6`, `-5`, `-4`, `-3`, `-2`, `-1`, `0`, `1`, `2`, `3`, `4`, `5`, `6`)
```

(ref:descriptives-liss-cap) Mean and Standard Deviation of the Big Five and Life Satisfaction over Time in the LISS Panel.

(ref:descriptives-note) Standard deviation shown in brackets; $time=0$ marks the first year where the transition to grandparenthood has been reported.

```{r descriptives-liss, results="asis", warning=FALSE, message=FALSE}
# no duplicate rownames allowed :(
rownames(descriptives_liss) <- c("Grandparents \\textcolor{white}{A}",
                          "\\textcolor{white}{Ag}", 
                          "Parent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{Ap}", 
                          "Nonparent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{An}", 
                          "Grandparents \\textcolor{white}{C}",
                          "\\textcolor{white}{Cg}", 
                          "Parent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cp}", 
                          "Nonparent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cn}", 
                          "Grandparents \\textcolor{white}{E}",
                          "\\textcolor{white}{Eg}", 
                          "Parent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{Ep}", 
                          "Nonparent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{En}", 
                          "Grandparents \\textcolor{white}{N}",
                          "\\textcolor{white}{Ng}", 
                          "Parent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Np}", 
                          "Nonparent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Nn}", 
                          "Grandparents \\textcolor{white}{O}",
                          "\\textcolor{white}{Og}", 
                          "Parent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{Op}", 
                          "Nonparent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{On}", 
                          "Grandparents \\textcolor{white}{L}",
                          "\\textcolor{white}{Lg}", 
                          "Parent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Lp}", 
                          "Nonparent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Ln}")

apa_table(
  descriptives_liss
  , caption = "(ref:descriptives-liss-cap)"
  , note = "(ref:descriptives-note)"
  , align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
  , col.names = c("", "-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")
  , col_spanners = list("Pre-transition years" = c(2,7), "Post-transition years" = c(8,14))
  , stub_indents = list("Agreeableness" = 1:6, "Conscientiousness" = 7:12, 
                        "Extraversion" = 13:18, "Neuroticism" = 19:24,
                        "Openness" = 25:30, "Life satisfaction" = 31:36)
  , format.args = list(gt1 = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T),
                       digits = c(0,0,0,0,0,0,0,0,0,0,0,0,0),
                       zero = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T))
  , escape = F
  , landscape = T
  , placement = "ht"
  , font_size = "small"
)
```

(ref:descriptives-hrs-cap) Mean and Standard Deviation of the Big Five and Life Satisfaction over Time in the HRS.

```{r descriptives-hrs, results="asis", warning=FALSE, message=FALSE}
# no duplicate rownames allowed :(
rownames(descriptives_hrs) <- c("Grandparents \\textcolor{white}{A}",
                          "\\textcolor{white}{Ag}", 
                          "Parent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{Ap}", 
                          "Nonparent controls \\textcolor{white}{A}", 
                          "\\textcolor{white}{An}", 
                          "Grandparents \\textcolor{white}{C}",
                          "\\textcolor{white}{Cg}", 
                          "Parent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cp}", 
                          "Nonparent controls \\textcolor{white}{C}", 
                          "\\textcolor{white}{Cn}", 
                          "Grandparents \\textcolor{white}{E}",
                          "\\textcolor{white}{Eg}", 
                          "Parent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{Ep}", 
                          "Nonparent controls \\textcolor{white}{E}", 
                          "\\textcolor{white}{En}", 
                          "Grandparents \\textcolor{white}{N}",
                          "\\textcolor{white}{Ng}", 
                          "Parent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Np}", 
                          "Nonparent controls \\textcolor{white}{N}", 
                          "\\textcolor{white}{Nn}", 
                          "Grandparents \\textcolor{white}{O}",
                          "\\textcolor{white}{Og}", 
                          "Parent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{Op}", 
                          "Nonparent controls \\textcolor{white}{O}", 
                          "\\textcolor{white}{On}", 
                          "Grandparents \\textcolor{white}{L}",
                          "\\textcolor{white}{Lg}", 
                          "Parent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Lp}", 
                          "Nonparent controls \\textcolor{white}{L}", 
                          "\\textcolor{white}{Ln}")

apa_table(
  descriptives_hrs
  , caption = "(ref:descriptives-hrs-cap)"
  , note = "(ref:descriptives-note)"
  , align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
  , col.names = c("", "-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")
  , col_spanners = list("Pre-transition years" = c(2,7), "Post-transition years" = c(8,14))
  , stub_indents = list("Agreeableness" = 1:6, "Conscientiousness" = 7:12, 
                        "Extraversion" = 13:18, "Neuroticism" = 19:24,
                        "Openness" = 25:30, "Life satisfaction" = 31:36)
  , format.args = list(gt1 = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T),
                       digits = c(0,0,0,0,0,0,0,0,0,0,0,0,0),
                       zero = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T))
  , escape = F
  , landscape = T
  , placement = "ht"
  , font_size = "small"
)
```

(ref:stddiffmeans-balance-liss-cap) Standardized Difference in Means for Covariates Used in Propensity Score Matching and the Propensity Score in the LISS panel.

(ref:stddiffmeans-balance-note) PSM = propensity score matching, ref. = reference category, f. = female, m. = male, NA = covariate not used in this sample. The standardized difference in means between the grandparent and the two control groups (parent and nonparent) was computed by $(\bar{x}_{gp}-\bar{x}_{c})/ (\hat\sigma_{gp})$. Rules of thumb say that this measure should ideally be below $.25$ [@stuartMatchingMethodsCausal2010] or below $.10$ [@austinIntroductionPropensityScore2011].  

```{r stddiffmeans-balance-liss, results="asis"}
# Standardized differences between means
liss_balance_df_rownames <- liss_balance_df$Covariate
liss_balance_df <- liss_balance_df[-c(1)]
rownames(liss_balance_df) <- liss_balance_df_rownames

apa_table(
  liss_balance_df
  , caption = "(ref:stddiffmeans-balance-liss-cap)"
  , note = "(ref:stddiffmeans-balance-note)"
  , align = c("l", "l", "l", "r", "r", "r", "r")
  , col.names = c("Covariate", "Description", "Raw variable", 
                  "Before PSM", "After PSM", "Before PSM", "After PSM")
  , col_spanners = list("Parent control group" = c(4,5), "Nonparent control group" = c(6,7))
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  , escape = T
  , landscape = T
  , longtable = T
  , placement = "t"
  , font_size = "footnotesize"
)
```

(ref:stddiffmeans-balance-hrs-cap) Standardized Difference in Means for Covariates Used in Propensity Score Matching and the Propensity Score in the HRS.

```{r stddiffmeans-balance-hrs, results="asis"}
# Standardized differences between means
hrs_balance_df_rownames <- hrs_balance_df$Covariate
hrs_balance_df <- hrs_balance_df[-c(1)]
rownames(hrs_balance_df) <- hrs_balance_df_rownames

apa_table(
  hrs_balance_df
  , caption = "(ref:stddiffmeans-balance-hrs-cap)"
  , note = "(ref:stddiffmeans-balance-note)"
  , align = c("l", "l", "l", "r", "r", "r", "r")
  , col.names = c("Covariate", "Description", "Raw variable", 
                  "Before PSM", "After PSM", "Before PSM", "After PSM")
  , col_spanners = list("Parent control group" = c(4,5), "Nonparent control group" = c(6,7))
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  #, stub_indents = list("Basic demographics)" = 2:14, "Economic situation" = 15:22, "Health" = 23:30,
  #                      "Family and fertility" = 31:36, "Outcomes at the time of matching" = 37:42,
  #                      "Panel status" = 43:44)
  , escape = T
  , landscape = T
  , longtable = T
  , placement = "t"
  , font_size = "footnotesize"
)
```

(ref:H2-hetvar-tab-agree-cap) Tests of Heterogeneous Random Slope Variance Models for `r df_outcomes["agree", "maj"]` Against Comparison Models With a Uniform Random Slope Variance.

(ref:H2-hetvar-tab-note) The heterogeneous variance models (*df* = 16) differ only in the random effects from the comparison models (*df* = 13). In addition to two random slope variances (instead of one), the heterogeneous variance models estimate two additional random intercept/slope covariances. Both models estimate heterogeneous random intercept variances for the grandparent and control groups. *Var.* = random slope variance; *SD* = standard deviation; *LR* = likelihood ratio; *p* = p-value (of the LR test); *GP greater* = indicating if the random slope variance of the grandparents is larger than that of either control group.  

```{r H2-hetvar-tab-agree, results="asis"}
hetvar_agree_df <- rbind(
  cbind(hetvar_summaries$agree_lissanalysis_parents_hetvar, 
        hetvar_summaries$agree_lissanalysis_nonparents_hetvar),
  cbind(hetvar_summaries$agree_hrsanalysis_parents_hetvar, 
        hetvar_summaries$agree_hrsanalysis_nonparents_hetvar)
)
hetvar_agree_df[c(4,9)] <- lapply(hetvar_agree_df[c(4,9)], gsub, pattern="\\$p\\$", replacement="")
hetvar_agree_df[c(4,9)] <- lapply(hetvar_agree_df[c(4,9)], gsub, pattern="\\=", replacement="")
hetvar_agree_df[c(1:2, 4:5, 7:8, 10:11, 13:14, 16:17), c(3:5, 8:10)] <- ""

# make one large table
colnames(hetvar_agree_df) <- rep(c("Var.", "SD", "LR", "p", "GP greater"), 2)

# duplicate rownames are apparently forbidden - I circumvent this by adding a white letter to duplicates
rownames_hetvar_agree_df <- c("Before-slope: uniform \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "After-slope: uniform \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Shift: uniform \\textcolor{white}{L}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Before-slope: uniform \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "After-slope: uniform \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "Shift: uniform \\textcolor{white}{H}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{H}")
rownames(hetvar_agree_df) <- rownames_hetvar_agree_df

apa_table(
  hetvar_agree_df
  , caption = "(ref:H2-hetvar-tab-agree-cap)"
  , note = "(ref:H2-hetvar-tab-note)"
  , col.names = c("", 
                  "Var.", "SD", "LR", "p", "GP greater", 
                  "Var.", "SD", "LR", "p", "GP greater")
  , align = c("l", "r", "r", "r", "r", "c", "r", "r", "r", "r", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE),
                       digits = c(2,2,2,2,3,2,2,2,2,3,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE))
  , col_spanners = list("Parent controls" = c(2,6), "Nonparent controls" = c(7,11))
  #, stub_indents = list("Fixed effects" = 1:12)
  , escape = F
  , landscape = T
  , font_size = "scriptsize"
  , stub_indents = list("LISS" = 1:9, "HRS" = 10:18)
  , placement = "H"
)
```

(ref:H2-hetvar-tab-con-cap) Tests of Heterogeneous Random Slope Variance Models for `r df_outcomes["con", "maj"]` Against Comparison Models With a Uniform Random Slope Variance.

```{r H2-hetvar-tab-con, results="asis"}
hetvar_con_df <- rbind(
  cbind(hetvar_summaries$con_lissanalysis_parents_hetvar, 
        hetvar_summaries$con_lissanalysis_nonparents_hetvar),
  cbind(hetvar_summaries$con_hrsanalysis_parents_hetvar, 
        hetvar_summaries$con_hrsanalysis_nonparents_hetvar)
)
hetvar_con_df[c(4,9)] <- lapply(hetvar_con_df[c(4,9)], gsub, pattern="\\$p\\$", replacement="")
hetvar_con_df[c(4,9)] <- lapply(hetvar_con_df[c(4,9)], gsub, pattern="\\=", replacement="")
hetvar_con_df[c(1:2, 4:5, 7:8, 10:11, 13:14, 16:17), c(3:5, 8:10)] <- ""

# make one large table
colnames(hetvar_con_df) <- rep(c("Var.", "SD", "LR", "p", "GP greater"), 2)

# duplicate rownames are apparently forbidden - I circumvent this by adding a white letter to duplicates
rownames_hetvar_con_df <- c("Before-slope: uniform \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "After-slope: uniform \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Shift: uniform \\textcolor{white}{L}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Before-slope: uniform \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "After-slope: uniform \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "Shift: uniform \\textcolor{white}{H}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{H}")
rownames(hetvar_con_df) <- rownames_hetvar_con_df

apa_table(
  hetvar_con_df
  , caption = "(ref:H2-hetvar-tab-con-cap)"
  , note = "(ref:H2-hetvar-tab-note)"
  , col.names = c("", 
                  "Var.", "SD", "LR", "p", "GP greater", 
                  "Var.", "SD", "LR", "p", "GP greater")
  , align = c("l", "r", "r", "r", "r", "c", "r", "r", "r", "r", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE),
                       digits = c(2,2,2,2,3,2,2,2,2,3,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE))
  , col_spanners = list("Parent controls" = c(2,6), "Nonparent controls" = c(7,11))
  #, stub_indents = list("Fixed effects" = 1:12)
  , escape = F
  , landscape = T
  , font_size = "scriptsize"
  , stub_indents = list("LISS" = 1:9, "HRS" = 10:18)
  , placement = "H"
)
```

(ref:H2-hetvar-tab-extra-cap) Tests of Heterogeneous Random Slope Variance Models for `r df_outcomes["extra", "maj"]` Against Comparison Models With a Uniform Random Slope Variance.

```{r H2-hetvar-tab-extra, results="asis"}
hetvar_extra_df <- rbind(
  cbind(hetvar_summaries$extra_lissanalysis_parents_hetvar, 
        hetvar_summaries$extra_lissanalysis_nonparents_hetvar),
  cbind(hetvar_summaries$extra_hrsanalysis_parents_hetvar, 
        hetvar_summaries$extra_hrsanalysis_nonparents_hetvar)
)
hetvar_extra_df[c(4,9)] <- lapply(hetvar_extra_df[c(4,9)], gsub, pattern="\\$p\\$", replacement="")
hetvar_extra_df[c(4,9)] <- lapply(hetvar_extra_df[c(4,9)], gsub, pattern="\\=", replacement="")
hetvar_extra_df[c(1:2, 4:5, 7:8, 10:11, 13:14, 16:17), c(3:5, 8:10)] <- ""

# make one large table
colnames(hetvar_extra_df) <- rep(c("Var.", "SD", "LR", "p", "GP greater"), 2)

# duplicate rownames are apparently forbidden - I circumvent this by adding a white letter to duplicates
rownames_hetvar_extra_df <- c("Before-slope: uniform \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "After-slope: uniform \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Shift: uniform \\textcolor{white}{L}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Before-slope: uniform \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "After-slope: uniform \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "Shift: uniform \\textcolor{white}{H}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{H}")
rownames(hetvar_extra_df) <- rownames_hetvar_extra_df

apa_table(
  hetvar_extra_df
  , caption = "(ref:H2-hetvar-tab-extra-cap)"
  , note = "(ref:H2-hetvar-tab-note)"
  , col.names = c("", 
                  "Var.", "SD", "LR", "p", "GP greater", 
                  "Var.", "SD", "LR", "p", "GP greater")
  , align = c("l", "r", "r", "r", "r", "c", "r", "r", "r", "r", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE),
                       digits = c(2,2,2,2,3,2,2,2,2,3,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE))
  , col_spanners = list("Parent controls" = c(2,6), "Nonparent controls" = c(7,11))
  #, stub_indents = list("Fixed effects" = 1:12)
  , escape = F
  , landscape = T
  , font_size = "scriptsize"
  , stub_indents = list("LISS" = 1:9, "HRS" = 10:18)
  , placement = "H"
)
```

(ref:H2-hetvar-tab-neur-cap) Tests of Heterogeneous Random Slope Variance Models for `r df_outcomes["neur", "maj"]` Against Comparison Models With a Uniform Random Slope Variance.

```{r H2-hetvar-tab-neur, results="asis"}
hetvar_neur_df <- rbind(
  cbind(hetvar_summaries$neur_lissanalysis_parents_hetvar, 
        hetvar_summaries$neur_lissanalysis_nonparents_hetvar),
  cbind(hetvar_summaries$neur_hrsanalysis_parents_hetvar, 
        hetvar_summaries$neur_hrsanalysis_nonparents_hetvar)
)
hetvar_neur_df[c(4,9)] <- lapply(hetvar_neur_df[c(4,9)], gsub, pattern="\\$p\\$", replacement="")
hetvar_neur_df[c(4,9)] <- lapply(hetvar_neur_df[c(4,9)], gsub, pattern="\\=", replacement="")
hetvar_neur_df[c(1:2, 4:5, 7:8, 10:11, 13:14, 16:17), c(3:5, 8:10)] <- ""

# make one large table
colnames(hetvar_neur_df) <- rep(c("Var.", "SD", "LR", "p", "GP greater"), 2)

# duplicate rownames are apparently forbidden - I circumvent this by adding a white letter to duplicates
rownames_hetvar_neur_df <- c("Before-slope: uniform \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "After-slope: uniform \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Shift: uniform \\textcolor{white}{L}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Before-slope: uniform \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "After-slope: uniform \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "Shift: uniform \\textcolor{white}{H}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{H}")
rownames(hetvar_neur_df) <- rownames_hetvar_neur_df

apa_table(
  hetvar_neur_df
  , caption = "(ref:H2-hetvar-tab-neur-cap)"
  , note = "(ref:H2-hetvar-tab-note)"
  , col.names = c("", 
                  "Var.", "SD", "LR", "p", "GP greater", 
                  "Var.", "SD", "LR", "p", "GP greater")
  , align = c("l", "r", "r", "r", "r", "c", "r", "r", "r", "r", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE),
                       digits = c(2,2,2,2,3,2,2,2,2,3,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE))
  , col_spanners = list("Parent controls" = c(2,6), "Nonparent controls" = c(7,11))
  #, stub_indents = list("Fixed effects" = 1:12)
  , escape = F
  , landscape = T
  , font_size = "scriptsize"
  , stub_indents = list("LISS" = 1:9, "HRS" = 10:18)
  , placement = "H"
)
```

(ref:H2-hetvar-tab-open-cap) Tests of Heterogeneous Random Slope Variance Models for `r df_outcomes["open", "maj"]` Against Comparison Models With a Uniform Random Slope Variance.

```{r H2-hetvar-tab-open, results="asis"}
hetvar_open_df <- rbind(
  cbind(hetvar_summaries$open_lissanalysis_parents_hetvar, 
        hetvar_summaries$open_lissanalysis_nonparents_hetvar),
  cbind(hetvar_summaries$open_hrsanalysis_parents_hetvar, 
        hetvar_summaries$open_hrsanalysis_nonparents_hetvar)
)
hetvar_open_df[c(4,9)] <- lapply(hetvar_open_df[c(4,9)], gsub, pattern="\\$p\\$", replacement="")
hetvar_open_df[c(4,9)] <- lapply(hetvar_open_df[c(4,9)], gsub, pattern="\\=", replacement="")
hetvar_open_df[c(1:2, 4:5, 7:8, 10:11, 13:14, 16:17), c(3:5, 8:10)] <- ""

# make one large table
colnames(hetvar_open_df) <- rep(c("Var.", "SD", "LR", "p", "GP greater"), 2)

# duplicate rownames are apparently forbidden - I circumvent this by adding a white letter to duplicates
rownames_hetvar_open_df <- c("Before-slope: uniform \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "After-slope: uniform \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Shift: uniform \\textcolor{white}{L}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Before-slope: uniform \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "After-slope: uniform \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "Shift: uniform \\textcolor{white}{H}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{H}")
rownames(hetvar_open_df) <- rownames_hetvar_open_df

apa_table(
  hetvar_open_df
  , caption = "(ref:H2-hetvar-tab-open-cap)"
  , note = "(ref:H2-hetvar-tab-note)"
  , col.names = c("", 
                  "Var.", "SD", "LR", "p", "GP greater", 
                  "Var.", "SD", "LR", "p", "GP greater")
  , align = c("l", "r", "r", "r", "r", "c", "r", "r", "r", "r", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE),
                       digits = c(2,2,2,2,3,2,2,2,2,3,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE))
  , col_spanners = list("Parent controls" = c(2,6), "Nonparent controls" = c(7,11))
  #, stub_indents = list("Fixed effects" = 1:12)
  , escape = F
  , landscape = T
  , font_size = "scriptsize"
  , stub_indents = list("LISS" = 1:9, "HRS" = 10:18)
  , placement = "H"
)
```

(ref:H2-hetvar-tab-swls-cap) Tests of Heterogeneous Random Slope Variance Models for `r df_outcomes["swls", "maj"]` Against Comparison Models With a Uniform Random Slope Variance.

```{r H2-hetvar-tab-swls, results="asis"}
hetvar_swls_df <- rbind(
  cbind(hetvar_summaries$swls_lissanalysis_parents_hetvar, 
        hetvar_summaries$swls_lissanalysis_nonparents_hetvar),
  cbind(hetvar_summaries$swls_hrsanalysis_parents_hetvar, 
        hetvar_summaries$swls_hrsanalysis_nonparents_hetvar)
)
hetvar_swls_df[c(4,9)] <- lapply(hetvar_swls_df[c(4,9)], gsub, pattern="\\$p\\$", replacement="")
hetvar_swls_df[c(4,9)] <- lapply(hetvar_swls_df[c(4,9)], gsub, pattern="\\=", replacement="")
hetvar_swls_df[c(1:2, 4:5, 7:8, 10:11, 13:14, 16:17), c(3:5, 8:10)] <- ""

# make one large table
colnames(hetvar_swls_df) <- rep(c("Var.", "SD", "LR", "p", "GP greater"), 2)

# duplicate rownames are apparently forbidden - I circumvent this by adding a white letter to duplicates
rownames_hetvar_swls_df <- c("Before-slope: uniform \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "After-slope: uniform \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{L}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Shift: uniform \\textcolor{white}{L}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{L}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{L}", 
                              "Before-slope: uniform \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Before-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "After-slope: uniform \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (controls) \\textcolor{white}{H}", 
                              "After-slope: heterogeneous (grandparents) \\textcolor{white}{H}", 
                              "Shift: uniform \\textcolor{white}{H}", 
                              "Shift: heterogeneous (controls) \\textcolor{white}{H}", 
                              "Shift: heterogeneous (grandparents) \\textcolor{white}{H}")
rownames(hetvar_swls_df) <- rownames_hetvar_swls_df

apa_table(
  hetvar_swls_df
  , caption = "(ref:H2-hetvar-tab-swls-cap)"
  , note = "(ref:H2-hetvar-tab-note)"
  , col.names = c("", 
                  "Var.", "SD", "LR", "p", "GP greater", 
                  "Var.", "SD", "LR", "p", "GP greater")
  , align = c("l", "r", "r", "r", "r", "c", "r", "r", "r", "r", "c")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE),
                       digits = c(2,2,2,2,3,2,2,2,2,3,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, FALSE, TRUE))
  , col_spanners = list("Parent controls" = c(2,6), "Nonparent controls" = c(7,11))
  #, stub_indents = list("Fixed effects" = 1:12)
  , escape = F
  , landscape = T
  , font_size = "scriptsize"
  , stub_indents = list("LISS" = 1:9, "HRS" = 10:18)
  , placement = "H"
)
```


(ref:H3-rankorder-tab-cap) Rank-order stability.

(ref:H3-rankorder-tab-note) Correlations as indicators of rank-order stability and p-values of the interactions between each outcome at the last pre-transition assessment and the *grandparent* variable in regression models predicting the first available post-transition assessment. The average time lags in years between these two assessments are `r printnum(mean(lissanalysis_parents_rank$yr_lag))` ($SD$ = `r printnum(sd(lissanalysis_parents_rank$yr_lag))`) for the LISS parent sample, `r printnum(mean(lissanalysis_nonparents_rank$yr_lag))` ($SD$ = `r printnum(sd(lissanalysis_nonparents_rank$yr_lag))`) for the LISS nonparent sample, `r printnum(mean(hrsanalysis_parents_rank$yr_lag))` ($SD$ = `r printnum(sd(hrsanalysis_parents_rank$yr_lag))`) for the HRS parent sample, and `r printnum(mean(hrsanalysis_nonparents_rank$yr_lag))` ($SD$ = `r printnum(sd(hrsanalysis_nonparents_rank$yr_lag))`) for the HRS nonparent sample. $Cor_{all}$ = rank-order stability; $Cor_{GP}$ = rank-order stability (only grandparents); $Cor_{con}$ = rank-order stability (only controls).

```{r H3-rankorder-tab, results="asis"}
rank_order_df_ordered <- rbind(
  cbind(rank_order_df[seq(1, 24, 4), ], # LISS parents
        rank_order_df[seq(2, 24, 4), ]), # LISS nonparents
  cbind(rank_order_df[seq(3, 24, 4), ], # HRS parents
        rank_order_df[seq(4, 24, 4), ]) # HRS nonparents
)

# duplicate rownames are apparently forbidden - I circumvent this by adding a white letter to duplicates
rownames_rank_order_df <- c(paste(df_outcomes[outcomes, "maj"], "\\textcolor{white}{L}"),
                            paste(df_outcomes[outcomes, "maj"], "\\textcolor{white}{H}"))
rownames(rank_order_df_ordered) <- rownames_rank_order_df

apa_table(
  rank_order_df_ordered
  , caption = "(ref:H3-rankorder-tab-cap)"
  , note = "(ref:H3-rankorder-tab-note)"
  , col.names = c("Outcome", 
                  "$Cor_{all}$", "$Cor_{GP}$", "$Cor_{con}$", "$p$", 
                  "$Cor_{all}$", "$Cor_{GP}$", "$Cor_{con}$", "$p$")
  , align = c("l", "r", "r", "r", "r", "r", "r", "r", "r")
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, FALSE),
                       digits = c(2,2,2,2,3,2,2,2,3),
                       zero = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, TRUE, TRUE, FALSE))
  , col_spanners = list("Parent controls" = c(2,5), "Nonparent controls" = c(6,9))
  , escape = F
  , landscape = T
  , font_size = "small"
  , stub_indents = list("LISS" = 1:6, "HRS" = 7:12)
  , placement = "H"
)
```

\newpage

\noindent
**Supplemental Figures**  

\newpage

\noindent 
**Complete Software and Session Information**  

We used `r cite_r("references-r.bib")` for data wrangling, analyses, and plots.  
The following is the output of R's *sessionInfo()* command, which shows information to aid analytic reproducibility of the analyses.

```{r session, results='asis'}

sessionInfo()

```

\newpage

\noindent 
**References**

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup
