---
title             : "The Transition to Grandparenthood and its Impact on the Big Five Personality Traits and Life Satisfaction"
shorttitle        : "Grandparenthood, Big Five, and Life Satisfaction"



author: 
  - name          : "Michael D. Kr&auml;mer" # I've noticed that Umlaute can pose a problem when knitting on Windows 
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "German Institute for Economic Research, Mohrenstr. 58, 10117 Berlin, Germany"
    email         : "mkraemer@diw.de"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Data Curation
      - Formal Analysis
      - Methodology
      - Visualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Manon A. van Scheppingen"
    affiliation   : "3"
    role:
      - Methodology
      - Writing - Review & Editing
  - name          : "William J. Chopik"
    affiliation   : "4"
    role:
      - Methodology
      - Writing - Review & Editing
  - name          : "David Richter"
    affiliation   : "1,4"
    role:
      - Supervision
      - Methodology
      - Writing - Review & Editing
      
affiliation:
  - id            : "1"
    institution   : "German Institute for Economic Research, Germany"
  - id            : "2"
    institution   : "International Max Planck Research School on the Life Course (LIFE), Max Planck Institute for Human Development, Germany"
  - id            : "3"
    institution   : "Tilburg University, Netherlands"
  - id            : "4"
    institution   : "Michigan State University, USA"
  - id            : "5"
    institution   : "Freie Universit&auml;t Berlin, Germany"

note: "\\clearpage"

authornote: |
  \addORCIDlink{Michael D. Krämer}{0000-0002-9883-5676}, Socio-Economic Panel (SOEP), German Institute for Economic Research (DIW Berlin); International Max Planck Research School on the Life Course (LIFE), Max Planck Institute for Human Development  
  Manon A. van Scheppingen, Department of Developmental Psychology, Tilburg School of Social and Behavioral Sciences, Tilburg University  
  William J. Chopik, Department of Psychology, Michigan State University  
  David Richter, Socio-Economic Panel (SOEP), German Institute for Economic Research (DIW Berlin); Survey Research Division, Department of Education and Psychology, Freie Universit&auml;t Berlin  

abstract: |
 abc  

keywords          : "grandparenthood, Big Five, life satisfaction, development, propensity score matching"
wordcount         : "abc"

bibliography      : ["references-zotero.bib", "references-r.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

csl               : "apa.csl"
documentclass     : "apa7"
classoption       : "man, noextraspace"
output            : papaja::apa6_pdf

header-includes:
  - \usepackage{setspace}
  - \usepackage{amsmath}
#  - \usepackage{float} # these two prevent some weird floating behavior of the figures
#  - \floatplacement{figure}{H}
  - \AtBeginEnvironment{tabular}{\singlespacing}
  - \AtBeginEnvironment{lltable}{\singlespacing}
  - \AtBeginEnvironment{tablenotes}{\doublespacing}
  - \captionsetup[table]{font={stretch=1.5}}
  - \captionsetup[figure]{font={stretch=1.5}}
#  - \raggedbottom # prevents LaTeX from varying the spacing between paragraphs to minimise empty space on pages
#  - \usepackage{floatrow}  # https://github.com/crsh/papaja/issues/342
#  - \floatsetup[figure]{capposition=top}

#appendix: gp-manuscript-papaja-appendix.Rmd

---

```{r setup, include = FALSE}
library("papaja") # not on CRAN, see https://github.com/crsh/papaja
library("citr")
library("tidyverse")
#library("psych") # for 'describe' demographics
#library("GPArotation") # needed for omega function in psych
#library("robustlmm") # for robust estimation (appendix)
#library("lme4")
#library("lmerTest")
#library("interactions") # for plotting interactions
#library("jtools") # for plot config
#library("cowplot") # for plot grid
#library("effects")
#library("patchwork") # for plot grid
#library("scales")
#library("car")
#library("careless") 
#library("corrplot") # appendix
#library("magick") # needed for figure cropping
#library("png") # appendix: external png file
#library("grid") # also needed for external png file because knitr::include_graphics() is buggy
r_refs("references-r.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)

# Set default chunk options (can be overridden in later chunks)
knitr::opts_chunk$set(echo = FALSE, 
                      eval = TRUE,
                      message = FALSE, 
                      fig.path='Figs/',
                      fig.width = 7, 
                      fig.height = 7,
                      results = "hide",
                      error = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      cache = FALSE)
```

```{r load data, include=FALSE}
#setwd("")

# Instead of specifying setwd() in the script, choose the working directory from the console after setting option in Rstudio:
# Tools -> Global Options -> R Markdown -> Evaluate chunks in directory
# The default working directory is the directory of the Rmd file (option "Document"). 
# Change this option to the current working directory of your R console (option "Current"), and then change
# your working directory via the console to the location where all the files are located.

# If you don't see this option, then execute install.packages("rmarkdown), close RStudio, and open again.

#source("") # This executes data import & cleaning.

```
In view of an aging demographic and an increased share of childcare functions being fulfilled by grandparents, intergenerational relations have received heightened attention from psychological and sociological research in recent years [@bengtsonNuclearFamilyIncreasing2001].<!-- weak 1st sentence --> With regard to personality development, the transition to grandparenthood has been posited as an important developmental task in old age [@huttemanDevelopmentalTasksFramework2014]. However, empirical research into the psychological consequences of this transition is sparse. Testing hypotheses derived from neo-socioanalytic theory [@robertsPersonalityDevelopmentContext2006] in a matched control-group design [see @luhmannStudyingChangesLife2014], we aim to investigate whether the transition to grandparenthood affects the Big Five personality traits and life satisfaction<!-- add a sentence on the structure of WB -->.  
<!-- introduce Big Five -->

## Personality Development in Middle Adulthood and Old Age
<!-- ~30–60 years / ~60 years and older -->
In accordance with the life span perspective characterizing aging as a lifelong process of development and adaptation [@baltesLifeSpanTheory2006], personality traits are subject to change throughout the entire life span [@spechtWhatDrivesAdult2014; @spechtPersonalityDevelopmentAdulthood2017; @costaPersonalityLifeSpan2019]. Although a major portion of development takes place in adolescence and emerging adulthood [@schwabaIndividualDifferencesPersonality2018; @bleidornPersonalityDevelopmentEmerging2017], evidence has accumulated that the Big Five personality traits also undergo changes in middle and old adulthood [e.g., @kandlerPatternsSourcesPersonality2015a; @lucasPersonalityDevelopmentLife2011; @mottusPersonalityTraitsOld2012; @wagnerPersonalityTraitDevelopment2016; for a review, see @spechtPersonalityDevelopmentAdulthood2017].  
Changes over time occur both in mean trait levels of these age groups [i.e., mean-level change; @robertsPatternsMeanlevelChange2006a] and in the relative ordering of people to each other on trait dimensions [i.e., rank-order stability; @anusicStabilityChangePersonality2016; @robertsRankorderConsistencyPersonality2000]<!-- distinction not very easy to follow... -->. Mean-level changes in middle adulthood [ca. 30--60 years old; @huttemanDevelopmentalTasksFramework2014] are typically characterized in terms of greater maturity as evidenced by increased agreeableness and conscientiousness, and decreased neuroticism [@robertsPatternsMeanlevelChange2006a]. In old age [ca. 60 years and older; @huttemanDevelopmentalTasksFramework2014], research is generally more sparse but there is some evidence for a reversal of the maturity effect, especially following retirement [sometimes termed *La dolce vita* effect; @marshMeasurementInvarianceBigfive2013; cf. @schwabaPersonalityTraitDevelopment2019] and at the end of life in ill health [@wagnerPersonalityTraitDevelopment2016]. In terms of rank-order stability, some prior studies have shown support for an inverted U-shape trajectory [@ardeltStillStableAll2000; @lucasPersonalityDevelopmentLife2011; @spechtStabilityChangePersonality2011; @wortmanStabilityChangeBig2012]: Rank-order stability rises until reaching a plateau in midlife, and decreases, again, in old age. However, evidence is mixed whether rank-order stability actually decreases again in old age [see @costaPersonalityLifeSpan2019]. Nonetheless, the historical view that personality is stable, or "set like plaster" [@spechtPersonalityDevelopmentAdulthood2017, p. 64] after one reaches adulthood [or leaves emerging adulthood behind; @bleidornPersonalityDevelopmentEmerging2017] can be largely abandoned [@spechtWhatDrivesAdult2014]<!-- can it? better ref? -->.  
Theories explaining the mechanisms of personality development in middle adulthood and old age emphasize as interdependent sources of stability and change both genetic influences and life experiences [@wagnerIntegrativeModelSources2020; @spechtWhatDrivesAdult2014]. Here, we focus on the latter[^f1] and conceptualize the transition to grandparenthood as a life experience that offers the adoption of a new social role according to the social investment principle of neo-socioanalytic theory [@robertsPersonalityDevelopmentContext2006; @lodi-smithSocialInvestmentPersonality2007]. According to the social investment principle, normative life events or transitions such as entering the work force or becoming a parent lead to personality maturation through the adoption of new social roles [@robertsEvaluatingFiveFactor2005]. These new roles encourage or compel people to act in a more agreeable, conscientious, and emotionally stable way, and the experiences in these role as well as societal expectations towardds them are hypothesized to drive long-term personality development [@lodi-smithSocialInvestmentPersonality2007]. Conversely, consistent social roles foster personality stability. The paradoxical theory of personality coherence [@caspiWhenIndividualDifferences1993] offers another explanation for personality development stating that trait change is more likely whenever people transition into unknown environments where pre-existing behavioral responses are no longer appropriate and societal norms or social expectations give clear indications how to behave instead (vs. environments where no such guidance is available). This supports the view that age-graded, normative life experiences such as possibly the transition to grandparenthood drive personality development [see also @spechtWhatDrivesAdult2014].  
Certain life events such as the first romantic relationship [@wagnerFirstPartnershipExperience2015] or the transition from high school to university [@ludtkeRandomWalkUniversity2011] have (partly) been found to be accompanied by mean-level increases in line with the social investment principle [for a review, see @bleidornLifeEventsPersonality2018]. However, recent evidence regarding the transition to parenthood failed to empirically support the social investment principle [@asselmannTestingSocialInvestment2020; @vanscheppingenPersonalityTraitDevelopment2016]. An analysis of monthly trajectories of the Big Five before and after nine major life events only found limited support for the social investment principle, that is, small increases were only found in emotional stability following the transition to employment but not for the other traits or for the other life events theoretically linked to social investment [@denissenTransactionsLifeEvents2019]. It has also been emphasized recently that effects of life events on the Big Five personality trends generally tend to be small, and need to be properly analyzed using robust, prospective designs and appropriate control groups [@bleidornLifeEventsPersonality2018; @luhmannStudyingChangesLife2014].  
Overall, much remains unknown regarding the environmental factors underlying personality development in middle adulthood and old age. One indication that age-graded, normative life experiences contribute to change following a period of relative stability is recent research on retirement [@bleidornRetirementAssociatedChange2018; @schwabaPersonalityTraitDevelopment2019]. While these results were only partly in line with the social investment principle in terms of mean-level changes and displayed substantial individual differences in change trajectories, the authors also discuss that as social role "divestment" [@schwabaPersonalityTraitDevelopment2019, p. X] retirement functions differently compared to social investment which adds a role. The transition to grandparenthood could represent such an investment in older adulthood---given that grandparents have regular contact with their grandchild and actively take part in childcare [i.e., invest psychologically in the new grandparent role; @lodi-smithSocialInvestmentPersonality2007], to some degree.  

[^f1]: In a behavior-genetic twin study, Kandler et al. [-@kandlerPatternsSourcesPersonality2015a] found that environmental factors were the main source of personality development in old age<!-- check again -->.

## Grandparenthood

The transition to grandparenthood, that is, the birth of the first grandchild, can be described as a time-discrete life event marking the beginning of one's status as a grandparent [@luhmannSubjectiveWellbeingAdaptation2012]. In terms of characteristics of major life events [@luhmannDimensionalTaxonomyPerceived2020], the transition to grandparenthood stands out in that it is externally caused (by one's own children), while at the same time predictable (as soon as one's children reveal their family planning or pregnancy), as well as generally positive in valence and emotionally significant.  
Grandparenthood can also be characterized as a developmental task [@huttemanDevelopmentalTasksFramework2014] mostly associated with the period of (early) old age---although considerable variation in the age at the transition to grandparenthood exists both within and across cultures [@leopoldDemographyGrandparenthoodInternational2015; @skopekWhoBecomesGrandparent2017]. Still, the period where parents on average experience the birth of their first grandchild coincides with the end of midlife stability in terms of personality development [@spechtPersonalityDevelopmentAdulthood2017], where retirement, shifting social roles, and initial cognitive and health declines can potentially be disruptive to life circumstances putting personality development into motion [e.g., @muellerPersonalityDevelopmentOld2016; @stephanPhysicalActivityPersonality2014]. As a developmental task, grandparenthood is expected to follow a normative sequence of aging that is subject to societal expectations and values differing across cultures and historical time [@huttemanDevelopmentalTasksFramework2014].<!-- add something about shifting nature of grandparenthood? --> Mastering developmental tasks to a high degree is hypothesized to drive personality development towards maturation similarly to propositions by the social investment principle, that is, leading to higher levels of agreeableness and conscientiousness, and lower levels of neuroticism [@robertsEvaluatingFiveFactor2005; @robertsPersonalityDevelopmentContext2006]. In comparison to the transition to parenthood which has been found to be ambivalent in terms of both personality maturation and life satisfaction [@kramerImpactHavingChildren2020; @vanscheppingenPersonalityTraitDevelopment2016], Hutteman et al. [-@huttemanDevelopmentalTasksFramework2014] hypothesize that the transition to grandparenthood is generally seen as positive because it (usually) does not impose the stressful daily demands of childcare on grandparents.  
While we could not find prior studies investigating development of the Big Five over the transition to grandparenthood, there is some evidence on life satisfaction although it is conflicting: Past research on associations of grandparenthood with life satisfaction often relied on cross-sectional designs [e.g., @mahneGrandparenthoodSubjectiveWellBeing2014; @triadoGrandparentsWhoProvide2014]. Longitudinal studies utilizing panel data from the Survey of Health, Ageing and Retirement in Europe (SHARE) showed that the birth of a grandchild was followed by improvements to quality of life and life satisfaction only among women [@tanskanenTransitionGrandparenthoodSubjective2019], and only in first-time grandmothers via their daughters [@digessaBecomingGrandparentIts2019]. Several studies emphasized that grandparents actively involved in childcare experienced larger positive effects to life satisfaction [@arpinoGrandparentingEducationSubjective2018; @danielsbackaAssociationGrandparentalInvestment2016; @danielsbackaGrandparentalChildcareHealth2019]. On the other hand, fixed effects regression models[^f2] using SHARE data did not find any effects of first-time grandparenthood on life satisfaction regardless of grandparental investment and only minor decreases of grandmothers' depressive symptoms [@sheppardBecomingFirstTimeGrandparent2019].
In a similar vein, some prospective studies reported beneficial effects of the transition to grandparenthood and of grandparental childcare investment on various health measures, especially in women  [@chungLongitudinalEffectsGrandchild2018; @condonTransitionGrandparenthoodProspective2018; @digessaHealthImpactIntensive2016; @digessaImpactCaringGrandchildren2016]. Again, effects on self-rated health did not persevere in fixed effects analyses as reported in Ates [-@atesDoesGrandchildCare2017] who used longitudinal data from the German Aging Survey (DEAS).  <!-- add paragraph from demographic literature? -->

[^f2]: Fixed effects regression models exclusively rely on within-person variance [see @bruderlFixedEffectsPanelRegression2015; @mcneishFixedEffectsModels2019].  

## Current Study

Three research questions motivate the current study which is the first to analyze personality development over the transition to grandparenthood with regards to the Big Five traits:

1. What are the effects of the transition to grandparenthood on mean-level trajectories of the Big Five traits and life satisfaction?
2. How large are interindividual differences in intraindividual change for the Big Five traits and life satisfaction over the transition to grandparenthood?
3. How does the transition to grandparenthood affect rank-order stability of the Big Five traits and life satisfaction?

To address these questions, we will compare development over the transition to grandparenthood with that of matched participants that do not experience this transition during the study period [@luhmannStudyingChangesLife2014]. This is necessary because pre-existing differences in variables related to the development of the Big Five or life satisfaction between those who are observed to become a grandparent and those who are not introduce confounding bias when trying to estimate the effect of the transition to grandparenthood [e.g., @vanderweeleOutcomeWideLongitudinalDesigns2020]. Propensity score matching accounts for confounding through equating the groups in their propensity to experience the event in question, which is calculated from a broad range of covariates related to the event and the outcomes. Thereby, to address confounding balance between the covariates used to calculate the propensity score is also aimed for [@stuartMatchingMethodsCausal2010].  
We adopt a prospective design that tests effects of first-time grandparents against two propensity-score-matched control groups: first, a matched control group of parents (but not grandparents) with at least their oldest child in reproductive age, and, second, a matched control group of nonparents. This allows us to disentangle potential effects attributable to becoming a grandparent from effects attributable to being a parent, thus, addressing selection effects into grandparenthood and confounding more comprehensively than previous research. Our comparative design also controls for average age-related and historical trends in the Big Five traits and life satisfaction [@luhmannStudyingChangesLife2014], and enables us to report effects of the transition to grandparenthood unconfounded by instrumentation effects, which describe the tendency of reporting lower well-being scores with each repeated measurement [@bairdLifeSatisfactionLifespan2010]. We go beyond previous studies utilizing matched control groups [@anusicDoesPersonalityModerate2014; @anusicTestingSetpointTheory2014; @yapDoesPersonalityModerate2012] in that we performed the matching at a specific time point preceding the transition to grandparenthood (at least two years before) and not based on individual survey years. This design choice ensures that the covariates involved in the matching procedure are not already influenced by the event or anticipation of it [@greenlandQuantifyingBiasesCausal2003; @rosenbaumConsquencesAdjustmentConcomitant1984; @vanderweeleOutcomeWideLongitudinalDesigns2020; @vanderweelePrinciplesConfounderSelection2019; @elwertEndogenousSelectionBias2014], thereby also reducing the risk of confounding through collider bias [@elwertEndogenousSelectionBias2014]. Similar approaches in the study of life events have recently been adopted [@balboRoleFamilyOrientations2016; @vanscheppingenTrajectoriesLifeSatisfaction2020; @kramerImpactHavingChildren2020].  
Informed by the social investment principle and previous research on personality development in middle adulthood and old age, we preregistered the following hypotheses (prior to data analysis; osf.io/):

* H1a: Following the birth of their first grandchild, grandparents increase slightly in agreeableness and conscientiousness, and decrease in neuroticism as compared to the matched control groups of parents (but not grandparents) and nonparents, but do not differ in their trajectories of extraversion and openness to experience.
* H1b: Grandmothers increase in life satisfaction following the transition to grandparenthood as compared to the matched control groups (but grandfathers do not).
* H2: Individual differences in intraindividual change in the Big Five and life satisfaction are larger in the grandparent group than the control group.
* H3a: Compared to the matched control groups, grandparents' rank-order stability of the Big Five decreases over the transition to grandparenthood.
* H3b: Grandparents' rank-order stability of life satisfaction is comparatively stable over the transition to grandparenthood.

Exploratorily, we further probe the social investment principle by testing two moderators of potential social investment and role conflict, hours of grandchild care and performing paid work.

# Methods

## Samples

To evaluate these hypotheses, we used data from two population-representative panel studies: the Longitudinal Internet Studies for the Social Sciences (LISS) panel from the Netherlands and the Health and Retirement Study (HRS) from the United States.  
The LISS panel is a representative sample of the Dutch population initiated in 2008 with data collection still ongoing [@scherpenzeelDataCollectionProbabilityBased2011; @vanderlaanRepresentativityLISSPanel2009]. It is administered by CentERdata (Tilburg University, The Netherlands). Included households are a true probability sample of households drawn from the population register [@scherpenzeelTrueLongitudinalProbabilitybased2010]. While originally roughly half of invited households consented to participate, refreshment samples were drawn in order to oversample previously underrepresented groups using information about response rates and their association with demographic variables (household type, age, ethnicity; see https://www.lissdata.nl/about-panel/sample-and-recruitment). Data collection was carried out online and participants lacking the necessary technical equipment were outfitted with it. We included yearly assessments from 2008 to 2019 from several different modules (see *Measures*) <!-- the so-called focal months for the yearly assessments differ both between modules and across years -->as well as data on basic demographics which was assessed on a monthly rate. For later coding of covariates from these monthly demographic data we used the first available assessment in each year.  
The HRS is a longitudinal population-representative study of older adults in the US [@sonnegaCohortProfileHealth2014] administered by the Survey Research Center (University of Michigan, United States). Initiated in 1992 with a first cohort of individuals aged 51-61 and their spouses, the study has since been extended with additional cohorts in the 1990s. In addition to the HRS core interview every two years (in-person or as a telephone survey), the study has since 2006 included a leave-behind questionnaire covering a broad range of psychosocial topics including the Big Five personality traits and life satisfaction. These topics, however, were only administered every four years starting in 2006 for one half of the sample and in 2008 for the other half. We included personality data from 2006 to 2016, all available data for the coding of the transition to grandparenthood from 1996 to 2016, as well as covariate data from 2006 to 2016 including variables drawn from the Imputations File and the Family Data (available up to 2014).  
These two panel studies provided the advantage that they contained several waves of personality data as well as information on grandparent status and a broad range of covariates at each wave. While the HRS provided a large sample with a wider age range, the LISS panel was smaller and younger[^f3] but provided more frequent personality assessments spaced every one to two years. Note that M. van Scheppingen has previously used the LISS panel to analyze ???. B. Chopik has previously used the HRS to analyze ???. These publications do not overlap with the current study in the central focus of grandparenthood.[^f4] The present study used de-identified archival data in the public domain, and, thus, it was not necessary to obtain ethical approval from an IRB.

[^f3]: The reason for the included grandparents from the LISS panel being younger was that grandparenthood questions were part of the *Work and Schooling* module and---for reasons unknown to us---filtered to participants performing paid work. Thus, older, retired first-time grandparents from the LISS panel could not be included.

[^f4]: Publications using LISS panel data can be searched at https://www.dataarchive.lissdata.nl/publications/. Publications using HRS data can be searched at https://hrs.isr.umich.edu/publications/biblio/. 

## Measures

### Personality 

In the LISS panel, the Big Five personality traits were assessed using the 50-item version of the IPIP Big-Five inventory scales [@goldbergDevelopmentMarkersBigFive1992]. For each Big Five trait, ten 5-point Likert-scale items were answered (1 = *very inaccurate*, 2 = *moderately inaccurate*, 3 = *neither inaccurate nor accurate*, 4 = *moderately accurate*, 5 = *very accurate*). Example items included "Like order" (conscientiousness),  "Sympathize with others' feelings" (agreeableness), "Worry about things" (neuroticism), "Have a vivid imagination" (openness to experience), and "Start conversations" (extraversion). At each wave, we took a participant's mean of each subscale as their trait score. Internal consistencies, as indicated by McDonald's $\omega$ [@mcneishThanksCoefficientAlpha2018], averaged XX over all traits and years ranging from XX (X) in year to XX (X) in year. Another study has shown measurement invariance for these scales across time and age groups [@schwabaIndividualDifferencesPersonality2018]. The Big Five (and life satisfaction) were contained in the *Personality* module which was administered yearly but with planned missingness in some years for certain cohorts [see @denissenTransactionsLifeEvents2019]. Thus, there are one to two years between included assessments, given no other sources of missingness.  
In the HRS, the Midlife Development Inventory (MIDI) scales were administered to measure the Big Five [@lachmanMidlifeDevelopmentInventory1997]. This scale was constructed for use in large-scale panel studies of adults and consisted of 26 adjectives (five each for conscientiousness, agreeableness, and extraversion, four for neuroticism, and seven for openness to experience). Participants were asked to rate on a 4-point scale how well each item described them (1 = *a lot*, 2 = *some*, 3 = *a little*, 4 = *not at all*). Example items included "Organized" (conscientiousness), "Sympathetic" (agreeableness), "Worrying" (neuroticism), "Imaginative" (openness to experience), and "Talkative" (extraversion). For better comparability with the LISS panel, we reverse scored all items so that higher values corresponded to higher trait levels<!-- to be done --> and, at each wave, took the mean of each subscale as the trait score. Big Five trait scores showed satisfactory internal consistencies which averaged XX over all traits and years ranging from XX (X) in year to XX (X) in year.   

### Life satisfaction 

In both samples, life satisfaction was assessed using the 5-item Satisfaction with Life Scale [SWLS; @dienerSatisfactionLifeScale1985] which participants answered on a 7-point Likert scale (1 = *strongly disagree*, 2 = *somewhat disagree*, 3 = *slightly disagree*, 4 = *neither agree or disagree*, 5 = *slightly agree*, 6 = *somewhat agree*, 7 = *strongly agree*)[^f5]. An example item was "I am satisfied with my life". In the LISS panel, internal consistencies averaged XX over all years ranging from XX (X) in year to XX (X) in year. In the HRS, internal consistencies averaged XX over all years ranging from XX (X) in year to XX (X) in year.

[^f5]: In the LISS panel, the "somewhat" was omitted and instead of "or" "nor" was used.

### Transition to Grandparenthood

The procedure to obtain information on grandparents' transition to grandparenthood generally followed the same steps in both samples. The items this coding was based on, however, differed slightly: In the LISS panel, participants were asked "Do you have children and/or grandchildren?" with "children", "grandchildren", and "no children or grandchildren" as possible answer categories. This question was part of the *Work and Schooling* module and filtered to participants performing paid work. In the HRS, all participants were asked for the total number of grandchildren: "Altogether, how many grandchildren do you (or your husband / wife / partner, or your late husband / wife / partner) have? Include as grandchildren any children of your (or your [late] husband's / wife's / partner's) biological, step- or adopted children". [^f6]  
In both samples, we tracked grandparenthood status (0 = *no grandchildren*, 1 = *at least one grandchild*) over time. Due to longitudinally inconsistent data in some cases<!-- report N? -->, we included in the grandparent group only participants with exactly one transition from 0 to 1 in this grandparenthood status variable, and no transitions back (see Fig. SX). We marked participants who continually indicated that they had no grandchildren as potential members of the control groups.  

[^f6]: The reference to step- or adopted children has been added since wave 2006.

### Covariates

For propensity score matching, we used a broad set of covariates [@vanderweeleOutcomeWideLongitudinalDesigns2020] covering participants' demographics (e.g., education), economic situation (e.g., income), and health (e.g., mobility difficulties). We also included the pre-transition outcome variables as covariates---as recommended in the literature [@cookHowMuchBias2020; @hallbergPretestMeasuresStudy2018; @steinerImportanceCovariateSelection2010; @vanderweeleOutcomeWideLongitudinalDesigns2020], as well as the panel wave participation count and the assessment year in order to control for instrumentation effects and historical trends [e.g., 2008 financial crisis; @bairdLifeSatisfactionLifespan2010; @luhmannStudyingChangesLife2014]. For matching grandparents with the parent control group we additionally included as covariates variables related to fertility and family history (e.g., number of children, age of first three children) which were causally related to the timing of the transition to grandparenthood [i.e., entry into treatment; @arpinoFamilyHistoriesDemography2018; @margolisCohortPerspectiveDemography2019].  
Covariate selection has seldom been explicitly discussed in previous longitudinal studies estimating treatment effects of life events (e.g., through a matching design). We see two (in part conflicting) traditions that address covariate selection: First, classical recommendations from psychology argue to include all available variables that are to associated with both the treatment assignment process (i.e., selection into treatment) and the outcome [e.g., @steinerImportanceCovariateSelection2010; @stuartMatchingMethodsCausal2010]<!-- good refs? -->. Second, recommendations from a structural causal modeling perspective [see @elwertEndogenousSelectionBias2014; @rohrerThinkingClearlyCorrelations2018] are more cautious aiming to avoid pitfalls such as conditioning on a pre-treatment collider (collider bias) or amediator (overcontrol bias). Structural causal modeling, however, requires advanced knowledge of the causal structures underlying all involved variables [@pearlCausalInferenceStatistics2009]<!-- ref? Pearl? -->.  
In selecting covariates, we followed guidelines laid out by VanderWeele et al. [-@vanderweeleOutcomeWideLongitudinalDesigns2020; -@vanderweelePrinciplesConfounderSelection2019] which reconcile both views and offer practical guidance when complete knowledge of the underlying causal structures is unknown: They propose a "modified disjunctive cause criterion" [@vanderweelePrinciplesConfounderSelection2019, p. 218] recommending to select all available covariates which are assumed to be causes of the outcomes, treatment exposure (i.e., the transition to grandparenthood), or both, as well as any proxies for an unmeasured common cause of the outcomes and treatment exposure. To be excluded from this list are variables assumed to be instrumental variables (i.e., assumed causes of treatment exposure that are unrelated to the outcomes except through the exposure) and collider variables [@elwertEndogenousSelectionBias2014]. Because all our covariates were measured at the time of matching (i.e, at least two years before the birth of the grandchild), we judge the risk of covariates introducing collider bias and overcontrol bias to be relatively small.  
An overview of the variables we used to compute the propensity scores for matching can be found in the Supplemental Material<!-- or Table X? -->, alongside justification for each covariate on whether we assume it to be causally related to treatment assignment, the outcomes, or both. Generally, we tried to find substantively equivalent covariates in both samples but had to compromise in a few cases (e.g., children's educational level only in HRS vs. children living at home only in LISS).<!-- work in progress - justifying covariate selection -->  

### Moderators

Based on insights from previous research, we tested three variables as potential moderators of the mean-level trajectories<!-- only those or also everything else? --> of the Big Five and life satisfaction over the transition to grandparenthood: First, we analyzed whether gender acted as a moderator as indicated by research on life satisfaction. We coded a dummy variable indicating female gender (0 = *male*, 1 = *female*). Second, we tested whether performing paid work or not was associated with divergent trajectories of the Big Five and life satisfaction. Since the LISS panel subsample of grandparents we identified was based exclusively on participants performing paid work, we performed these analyses only in the HRS subsample. This served two purposes: to test how participants involved in the workforce (even if officially retired) differed from those not working <!-- which might shed light on role conflict -->, and to assess whether potential differences in the main results between the LISS and HRS samples disappeared once we constrained the HRS sample in the same way that the LISS sample had already been constrained through filtering. Third, we tested how the involvement in grandchild care affected trajectories of the Big Five and life satisfaction in grandparents after the transition to grandparenthood. We coded a dummy variable (0 = *provided less than 100 hours of grandchild care*, 1 = *provided 100 or more hours of grandchild care*) as a moderator based on the question "Did you (or your [late] husband / wife / partner) spend 100 or more hours in total since the last interview / in the last two years taking care of grand- or great grandchildren?". [^f7] This information was only available in the HRS; in the LISS panel only very few participants answered follow-up questions on intensity of care (less than 50 in the final analysis sample).

[^f7]: Although artificial dichotomization of a continuous variable (hours of care) is not ideal for moderation analysis, there were too many missing values in the variable assessing hours of care directly (*E063).

## Procedure

## Analytical Strategy


This design is referred to by Shadish, Cook, and Campbell [-@shadishExperimentalQuasiexperimentalDesigns2002, p.182] as an interrupted time-series with a "nonequivalent no-treatment control group".


use linear piecewise regression coefficients in multilevel regression models [@hoffmanLongitudinalAnalysisModeling2015] to analyze mean-level changes. Such piecewise growth-curve models have been used in similar analysis contexts [e.g., @bleidornRetirementAssociatedChange2018; @schwabaPersonalityTraitDevelopment2019; @vanscheppingenTrajectoriesLifeSatisfaction2020]. 

For this article, I want to focus on mean-level changes, but we can additionally analyze rank-order stability over the transition to grandparenthood as well as interindividual differences in intraindindivual changes 
Jaap Denissen modeled interindividual differences in personality trait change as variance in the slopes. Further, he modeled “more sudden shifts in rank-order trait differences” as the comparison of the test-retest correlations of the event and the control group [@denissenTransactionsLifeEvents2019]

A list of all software we used is provided in the Supplemental Material.  
    
# Results


# Discussion
Based on  

- personality maturation cross-culturally: [@bleidornPersonalityMaturationWorld2013; @chopikPersonalityChangeLife2018]

- facets / nuances [@mottusDevelopmentDetailsAge2021]

- arrival of grandchild associated with retirement decisions [@lumsdaineRetirementTimingWomen2015]

- Does the Transition to Grandparenthood Deter Gray Divorce? A Test of the Braking Hypothesis [@brownDoesTransitionGrandparenthood2021]

- prolonged period of grandparenthood? [@margolisHealthyGrandparenthoodHow2017]

- subjective experience of aging [@bordoneGrandchildrenInfluenceHow2015]

- policy relevance of personality [@bleidornPolicyRelevancePersonality2019], e.g., health outcomes [@turianoPersonalityTraitLevel2012], but not really evidence for healthy neuroticism [@turianoHealthyNeuroticismAssociated2020]

## Limitations

Despite  

## Conclusions
Our  

## Acknowledgements
We thank X for valuable feedback.  

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

