\renewcommand{\appendixname}{\textcolor{white}{.}}
\renewcommand{\thefigure}{S\arabic{figure}} \setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}} \setcounter{table}{0}
\setcounter{page}{1} <!-- to reset page numbering -->

# Supplemental Material

\noindent
**Supplemental Tables**  


(ref:piecewise-coding-scheme-cap) Coding scheme for the piecewise regression coefficients used to model the developmental trajectories around the transition to grandparenthood

(ref:piecewise-coding-scheme-note) $time=0$ marks the first year where the transition to grandparenthood has been reported.  

```{r piecewise-coding-scheme, results="asis"}
# Visualize the coding scheme for the piecewise regression coefficients
coding_scheme <- rbind(c(0, 1, 2, 3, 4, 5, 5, 5, 5, 5, 5, 5, 5),
                       c(0, 0, 0, 0, 0, 0, 1, 2, 3, 4, 5, 6, 7),
                       c(0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1, 1))
rownames(coding_scheme) <- c("Before-slope", "After-slope", "Jump")

apa_table(
  coding_scheme
  , caption = "(ref:piecewise-coding-scheme-cap)"
  , note = "(ref:piecewise-coding-scheme-note)"
  , align = c("l", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c", "c")
  , col.names = c("Coefficient", "-6", "-5", "-4", "-3", "-2", "-1", "0", "1", "2", "3", "4", "5", "6")
  , col_spanners = list("Pre-transition years" = c(2,7), "Post-transition years" = c(8,14))
  , format.args = list(gt1 = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T),
                       digits = c(0,0,0,0,0,0,0,0,0,0,0,0,0),
                       zero = c(T, T, T, T, T, T, T, T, T, T, T, T, T, T))
  , escape = T
  , landscape = F
  , placement = "ht"
  #, font_size = "footnotesize"
)
```

(ref:stddiffmeans-balance-liss-cap) Standardized Difference in Means for Covariates Used in Propensity Score Matching and the Propensity Score in the LISS 

(ref:stddiffmeans-balance-note) PSM = propensity score matching, ref. = reference category, f. = female, m. = male, NA = covariate not used in this sample. The standardized difference in means between the grandparent and the two control groups (parent and nonparent) was computed by $(\bar{x}_{gp}-\bar{x}_{c})/ (\hat\sigma_{gp})$. A rule of thumb says that this measure should ideally be below $.25$ [@stuartMatchingMethodsCausal2010].  

```{r stddiffmeans-balance-liss, results="asis"}
# Standardized differences between means
liss_balance_df_rownames <- liss_balance_df$Covariate
liss_balance_df <- liss_balance_df[-c(1)]
rownames(liss_balance_df) <- liss_balance_df_rownames

apa_table(
  liss_balance_df
  , caption = "(ref:stddiffmeans-balance-liss-cap)"
  , note = "(ref:stddiffmeans-balance-note)"
  , align = c("l", "l", "l", "r", "r", "r", "r")
  , col.names = c("Covariate", "Description", "Raw variable", 
                  "Before PSM", "After PSM", "Before PSM", "After PSM")
  , col_spanners = list("Parent control group" = c(4,5), "Nonparent control group" = c(6,7))
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  , escape = T
  , landscape = T
  , longtable = T
  , placement = "t"
  , font_size = "footnotesize"
)
```

(ref:stddiffmeans-balance-hrs-cap) Standardized Difference in Means for Covariates Used in Propensity Score Matching and the Propensity Score in the HRS 

```{r stddiffmeans-balance-hrs, results="asis"}
# Standardized differences between means
hrs_balance_df_rownames <- hrs_balance_df$Covariate
hrs_balance_df <- hrs_balance_df[-c(1)]
rownames(hrs_balance_df) <- hrs_balance_df_rownames

apa_table(
  hrs_balance_df
  , caption = "(ref:stddiffmeans-balance-hrs-cap)"
  , note = "(ref:stddiffmeans-balance-note)"
  , align = c("l", "l", "l", "r", "r", "r", "r")
  , col.names = c("Covariate", "Description", "Raw variable", 
                  "Before PSM", "After PSM", "Before PSM", "After PSM")
  , col_spanners = list("Parent control group" = c(4,5), "Nonparent control group" = c(6,7))
  , format.args = list(gt1 = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE),
                       digits = c(2,2,2,2,2,2,2),
                       zero = c(TRUE, TRUE, TRUE, TRUE, TRUE, TRUE, TRUE))
  #, stub_indents = list("Basic demographics)" = 2:14, "Economic situation" = 15:22, "Health" = 23:30,
  #                      "Family and fertility" = 31:36, "Outcomes at the time of matching" = 37:42,
  #                      "Panel status" = 43:44)
  , escape = T
  , landscape = T
  , longtable = T
  , placement = "t"
  , font_size = "footnotesize"
)
```

\newpage

\noindent
**Supplemental Figures**  

\newpage

\noindent 
**Complete Software and Session Information**  

We used `r cite_r("references-r.bib")` for data wrangling, analyses, and plots.  
The following is the output of R's *sessionInfo()* command, which shows information to aid analytic reproducibility of the analyses.

```{r session, results='asis'}

sessionInfo()

```

\newpage

\noindent 
**References**

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup
